---
title: "US Supply Chain Risk Analysis — Initial Data Analysis"
format: html
editor: visual
---

> Dataset background, source/licence, research questions, analysis plan and the risk discussion are in [Project_Proposal.qmd](Project_Proposal.qmd). This file holds the runnable initial data analysis only.

<https://www.kaggle.com/datasets/yuanchunhong/us-supply-chain-risk-analysis-dataset>

## Load data

```{r}
library(tidyverse)
library(lubridate)

supply_data <- read_csv("../Data/US_supply_chain_data.csv")
glimpse(supply_data)
```

```{r}
supply_data <- supply_data %>%
  mutate(
    Order_Date = ymd(Order_Date),
    Dispatch_Date = ymd(Dispatch_Date),
    Delivery_Date = ymd(Delivery_Date)
  )
nrow(supply_data)
ncol(supply_data)
```

## Summary statistics

```{r}
summary(supply_data$Delay_Days)
summary(supply_data$Supplier_Reliability_Score)
summary(supply_data$Historical_Disruption_Count)
summary(supply_data$Quantity_Ordered)
summary(supply_data$Order_Value_USD)
```

```{r}
# Count observations by shipping mode
supply_data %>%
  count(Shipping_Mode)

# Count observations by product category
supply_data %>%
  count(Product_Category)

# Count observations by supply risk flag
supply_data %>%
  count(Supply_Risk_Flag)
```

## Plot 1 — distribution of delivery delays

```{r}
ggplot(supply_data, aes(x = Delay_Days)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Delivery Delays",
    x = "Delay Days",
    y = "Number of Orders"
  ) +
  theme_minimal()
```

## Plot 2 — delivery delays by shipping mode

```{r}
ggplot(supply_data, aes(x = Shipping_Mode, y = Delay_Days)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Delivery Delays by Shipping Mode",
    x = "Shipping Mode",
    y = "Delay Days"
  ) +
  theme_minimal()
```

## Plot 3 — supplier reliability and delivery delays

```{r}
ggplot(supply_data, aes(x = Supplier_Reliability_Score, y = Delay_Days)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Supplier Reliability and Delivery Delays",
    x = "Supplier Reliability Score",
    y = "Delay Days"
  ) +
  theme_minimal()
```

## Plot 4 — historical disruptions and delays

```{r}
ggplot(supply_data, aes(x = Historical_Disruption_Count, y = Delay_Days)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Historical Disruptions and Delivery Delays",
    x = "Historical Disruption Count",
    y = "Delay Days"
  ) +
  theme_minimal()
```

## Predictive model (RQ3)

```{r}
delay_model <- lm(
  Delay_Days ~ Supplier_Reliability_Score +
    Historical_Disruption_Count +
    Shipping_Mode +
    Product_Category,
  data = supply_data
)

summary(delay_model)
```

## Correlation scan

```{r}
# Keep only numeric columns
num_data <- supply_data %>% select(where(is.numeric))

# Correlation matrix (pairwise to tolerate any NAs)
cor_mat <- cor(num_data, use = "pairwise.complete.obs")
round(cor_mat, 2)

# Keep only the upper triangle so it mirrors corrplot's type = "upper"
cor_upper <- cor_mat
cor_upper[lower.tri(cor_upper)] <- NA

# Reshape the matrix to long format (Var1, Var2, correlation)
cor_long <- as.data.frame(as.table(cor_upper)) %>%
  rename(corr = Freq) %>%
  filter(!is.na(corr))

ggplot(cor_long, aes(x = Var2, y = Var1, fill = corr)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", corr)), size = 2.5) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "white", high = "#2166AC",
    midpoint = 0, limits = c(-1, 1), name = "Correlation"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank()
  ) +
  labs(x = NULL, y = NULL, title = "Correlation matrix")
```

## Derived lead-time variables & group tests

```{r}
supply_data <- supply_data %>%
  mutate(
    days_to_dispatch  = as.numeric(Dispatch_Date - Order_Date),
    days_to_delivery  = as.numeric(Delivery_Date - Dispatch_Date),
    total_lead_time   = as.numeric(Delivery_Date - Order_Date)
  )

# Re-run the numeric correlation scan including these
supply_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs") %>%
  round(2)

# Does mean delay differ by category?
summary(aov(Delay_Days ~ Shipping_Mode, data = supply_data))
summary(aov(Delay_Days ~ Product_Category, data = supply_data))

# Group means to eyeball effect sizes
supply_data %>%
  group_by(Shipping_Mode) %>%
  summarise(mean_delay = mean(Delay_Days), n = n())
```

## Leakage check on Supply_Risk_Flag

```{r}
supply_data %>%
  group_by(Supply_Risk_Flag) %>%
  summarise(mean_delay = mean(Delay_Days),
            median_delay = median(Delay_Days),
            n = n())

# Flag is 1 exactly when Delay_Days > 0 -> derived from the outcome (leakage)
all(supply_data$Supply_Risk_Flag == as.integer(supply_data$Delay_Days > 0))
```

## Expanded model (leakage flag excluded)

```{r}
new_model <- lm(
  Delay_Days ~ Supplier_Reliability_Score + Historical_Disruption_Count +
    Shipping_Mode + Product_Category + Quantity_Ordered + Order_Value_USD,
  data = supply_data
)
summary(new_model)
```
