---
title: "Dataset Research"
format: html
editor: visual
---

# Dataset Research

## Lebensziele, Persönlichkeitsmerkmale und Erfolg von Studierenden verschiedener Fächergruppen

<https://search.gesis.org/research_data/SDN-10.7802-1763>

## Auswahlverfahren an deutschen Hochschulen

<https://search.gesis.org/research_data/SDN-10.7802-2292>

## Mental Health Care in the Last 4 Weeks, US only, possible Covid 19 impact questions

<https://catalog.data.gov/dataset/mental-health-care-in-the-last-4-weeks?from_hint=eyJzb3J0IjoicG9wdWxhcml0eSJ9>

```{r}
library(readr)
library(dplyr)
library(ggplot2)
data <- read_csv("Mental_Health_Care_in_the_Last_4_Weeks.csv")
head(data)
```

Each row reports one estimate: the percentage of US adults who, in the last 4 weeks, took a given mental-health-care action for a particular subgroup, geography and time period.

The key columns are:

-   `Indicator` (which action)
-   `Group`/`Subgroup` (which population is being described)
-   `State`
-   `Time Period` columns (when)
-   `Value` with its `LowCI`/`HighCI` confidence interval.

```{r}
print("== Overall shape ==")
dim(data)
print("== Indicators ==")
unique(data$Indicator)
print("== Population Dimensions ==")
unique(data$Group)
print("== How many time periods and geographies are covered ==")
length(unique(data$`Time Period`))
length(unique(data$State))
# Parse the start date into a real Date so we can plot over time
data <- data |>
  mutate(start_date = as.Date(`Time Period Start Date`, format = "%m/%d/%Y"))
range(data$start_date, na.rm = TRUE)
print("== Missing values: estimates suppressed for reliability reasons ==")
sum(is.na(data$Value))
```

```{r}
national <- data |>
  filter(Group == "National Estimate", State == "United States") |>
  filter(!is.na(Value))
ggplot(national, aes(x = start_date, y = Value, colour = Indicator)) +
  geom_line() +
  geom_point(size = 1) +
  labs(
    title = "US mental health care use, last 4 weeks (national estimate)",
    x = "Survey period (start date)",
    y = "Percent of adults",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 4))
```

```{r}
by_sex <- data |>
  filter(
    Group == "By Sex",
    State == "United States",
    Indicator == "Received Counseling or Therapy, Last 4 Weeks",
    !is.na(Value)
  )
ggplot(by_sex, aes(x = start_date, y = Value, colour = Subgroup)) +
  geom_line() +
  geom_point(size = 1) +
  labs(
    title = "Received counseling or therapy, by sex",
    x = "Survey period (start date)",
    y = "Percent of adults",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Source and licence

The data comes from the **U.S. Census Bureau's Household Pulse Survey**, published by the **National Center for Health Statistics (NCHS / CDC)** and distributed through [data.gov](https://catalog.data.gov/dataset/mental-health-care-in-the-last-4-weeks). The Household Pulse Survey is an experimental survey launched in 2020 to measure the social and economic impact of the COVID-19 pandemic on US households. As a work of the US federal government it is in the **public domain** (U.S. Government Work / no copyright), so it is free to use, redistribute and analyse with attribution to the source.

### What is in it

**Key variables:**

-   `Indicator` one of four actions (received counseling, needed but did not get counseling, took prescription medication, or medication and/or counseling).
-   `Group` / `Subgroup` how the population is split (Sex, Age, Race/Hispanic ethnicity, Education, Disability status, Sexual orientation, Gender identity, State, or the overall National Estimate).
-   `State` geography of the estimate.
-   `Time Period*` the survey window (number, label, start and end dates).
-   `Value`, `LowCI`, `HighCI` the estimated percentage and its 95% confidence interval.
-   `Suppression Flag` marks estimates withheld for being statistically unreliable.

### Research questions

1.  **Descriptive**
    -   Track each indicator's national value across the Aug 2020 – Dec 2021 window to see whether counseling, medication and unmet need rose, fell or stayed flat.
2.  **Exploratory**
    -   Compare values across sex, age, race/ethnicity and education to see which subgroups consistently report higher unmet need ("needed counseling but did not get it") or higher care use.
3.  **Exploratory/causal**
    -   Using the "By Presence of Symptoms of Anxiety/Depression" group, compare care use between those with and without symptoms.

# Data about dogs

needs access :( <https://data.dogagingproject.org/Index>
