---
title: "COVID-19 infection rates and unmet mental-health need"
author:
  - "Cedric Fernolend"
  - "Peiyue Liu"
date: today
---

We ask whether US states hit harder by COVID-19 also reported higher **unmet need for mental-health support**. The cleaning and joining happen in `code/01_download.R` → `code/02_clean.R`; this page loads the processed result and answers the two research questions.

```{r setup}
#| message: false
#| warning: false
library(tidyverse)
library(here)

joined   <- read_csv(here("data", "processed", "covid_mentalhealth_joined.csv"),
                     show_col_types = FALSE)
national <- read_csv(here("data", "processed", "covid_mentalhealth_national.csv"),
                     show_col_types = FALSE)

# National series used throughout RQ2
rq2 <- national |>
  arrange(win_start) |>
  filter(!is.na(unmet_need), !is.na(covid_rate))

rq2_diff <- rq2 |>
  mutate(d_need = unmet_need - lag(unmet_need),
         d_cov  = covid_rate - lag(covid_rate))
```

## RQ1: Does COVID intensity relate to unmet mental-health need?

**Question.** Across US states and survey windows, is a higher COVID-19 case rate
associated with a higher share of adults who needed counseling but did not
receive it?

The unit of analysis is one state in one survey window
(*n* = `r nrow(joined)`). The response is `unmet_need` (% of adults); the
predictor is `covid_rate` (new COVID cases per 100,000 residents accumulated over
the survey window). We estimate an ordinary least squares regression and report
the slope with its 95% confidence interval.

```{r rq1-model}
m1 <- lm(unmet_need ~ covid_rate, data = joined)
summary(m1)

# Population claim -> report the coefficient with its 95% CI, not just the p-value.
cbind(estimate = coef(m1), confint(m1))
```

```{r rq1-plot}
#| fig-cap: "Scatter plot of unmet mental-health need against COVID case rate. Each point is one US state in one Household Pulse survey window. The fitted ordinary least squares line is shown with its 95% confidence band."
#| fig-alt: "Scatter plot of percent of adults with unmet mental-health need on the y-axis against new COVID cases per 100k per survey window on the x-axis. Points form a wide cloud with a shallow upward-sloping regression line."
ggplot(joined, aes(x = covid_rate, y = unmet_need)) +
  geom_point(alpha = 0.3, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "red") +
  labs(
    title = "Higher COVID case rates track slightly higher unmet mental-health need",
    subtitle = "Weak but statistically significant positive association across states and windows",
    x = "New COVID cases per 100k during the survey window",
    y = "% adults who needed counseling but did not get it"
  ) +
  theme_minimal()
```

## RQ2: How did need and COVID co-move nationally over time?

**Question.** Following the country as a whole rather than comparing states, did
national unmet need and the national COVID case rate rise and fall together
between August 2020 and April 2022?

The unit of analysis is one survey window (*n* = `r nrow(rq2)`). We
describe the relationship graphically and with three statistics: the Pearson
product-moment and Spearman rank correlations in **levels**, and the Pearson
correlation of **first differences**, which removes any shared level or trend
that could otherwise generate a spurious level correlation.

```{r rq2-panels}
#| fig-cap: "Two-panel time series of national unmet mental-health need (top) and the national COVID-19 case rate (bottom), by survey window, Aug 2020 to Apr 2022, shown in their original units."
#| fig-alt: "Two stacked line charts sharing an x-axis of survey-window dates from Aug 2020 to Apr 2022. Top: percent of adults with unmet mental-health need. Bottom: new COVID cases per 100k, showing sharp winter peaks."
national |>
  select(win_start,
         `Unmet need (%)`    = unmet_need,
         `COVID cases /100k` = covid_rate) |>
  pivot_longer(-win_start) |>
  ggplot(aes(win_start, value)) +
  geom_line(colour = "darkblue") +
  geom_point(size = 1) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(title = "National unmet mental-health need and COVID case rate over time",
       x = "Survey window start", y = NULL) +
  theme_minimal()
```

```{r rq2-standardised}
#| fig-cap: "The same two national series, z-standardised so they share a common axis. Standardisation removes the differing units and scales, so vertical distance between the lines is directly interpretable as divergence in relative movement."
#| fig-alt: "Line chart with survey-window dates on the x-axis and standard deviations from the mean on the y-axis. COVID case rate shows tall sharp peaks in winter 2020-21 and winter 2021-22; unmet need stays within a narrow band near the mean throughout."
rq2 |>
  mutate(across(c(unmet_need, covid_rate), \(x) as.numeric(scale(x)))) |>
  select(win_start,
         `Unmet need`      = unmet_need,
         `COVID case rate` = covid_rate) |>
  pivot_longer(-win_start) |>
  ggplot(aes(win_start, value, colour = name)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1) +
  labs(
    title = "National unmet mental-health need and COVID case rate over time",
    subtitle = "Both series z-standardised to share one axis",
    x = "Survey window start",
    y = "Standard deviations from mean",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
Pearson product-moment and Spearman rank correlation, first differences: window-to-window changes
```{r rq2-metrics}
# Levels: Pearson product-moment and Spearman rank correlation
cor.test(rq2$unmet_need, rq2$covid_rate)
cor(rq2$unmet_need, rq2$covid_rate, method = "spearman")

# First differences: window-to-window changes
cor.test(rq2_diff$d_need, rq2_diff$d_cov)
```
Pearson product-moment and Spearman rank correlation, first differences: window-to-window changes (pre-omnicron)
```{r rq2-omicron}
# Is the relationship stable across the whole period, or does it break down
# during the Omicron surge of winter 2021-22?
pre_omicron <- rq2 |> filter(win_start < as.Date("2021-12-01"))

cor.test(pre_omicron$unmet_need, pre_omicron$covid_rate)

pre_omicron |>
  mutate(d_need = unmet_need - lag(unmet_need),
         d_cov  = covid_rate - lag(covid_rate)) |>
  filter(!is.na(d_need)) |>
  summarise(first_difference_r = cor(d_need, d_cov))
```

In **levels**, the association is moderate and positive: Pearson
*r* = 0.38 (95% CI [0.05, 0.64], *p* = 0.028), with a somewhat stronger
Spearman rank correlation of ρ = 0.53, indicating a monotonic rather than
strictly linear relationship.

In **first differences**, however, the association essentially vanishes:
*r* = 0.05 (*p* = 0.78). 

# Interpretation

## RQ1

The association is **positive, statistically significant, and substantively
weak**. The slope of 5.4 × 10⁻⁴ (95% CI [3.4 × 10⁻⁴, 7.5 × 10⁻⁴], *p* < 0.001)
implies that moving across the full observed range of case rates — from
essentially zero to roughly 5,000 cases per 100,000 in a single window —
corresponds to an increase in unmet need of only about **2.8 percentage points**,
against an outcome that itself ranges from roughly 3% to 21%. The model explains
approximately **1.6% of the variance** (R² = 0.016).

The narrow confidence interval and small *p*-value should therefore not be read as
evidence of a strong relationship. They largely reflect the **statistical power
conferred by a large sample** (*n* = 1,676 state-window observations): with this
many observations, even a negligible effect is estimated precisely enough to
exclude zero. Statistical significance and practical importance diverge sharply
here.

Because the observations are **state-level aggregates**, the estimate describes a
relationship between populations, not between individuals, and is exposed to the
**ecological fallacy** — the pattern across states need not hold for any
individual. The design is observational, so the coefficient is an association and
carries no causal interpretation.

## RQ2

In **levels**, the correlation is moderate and positive (Pearson
*r* = 0.38, 95% CI [0.05, 0.64], *p* = 0.028; Spearman ρ = 0.53, indicating a
monotonic rather than strictly linear relationship). In **first differences**,
the association essentially disappears (*r* = 0.05, *p* = 0.78).

That full-period summary, however, conceals a **structural break**. Restricting
the series to the first two pandemic waves (all windows before December 2021)
the level correlation nearly doubles to *r* = 0.72 (*n* = 27, *p* < 0.001). Across
2020 and most of 2021 the two series do track one another closely: unmet need
rose and fell broadly in step with the autumn 2020 and winter 2020–21 waves.

The relationship then **breaks down during the Omicron surge**. The window
beginning 29 December 2021 recorded roughly 2,504 cases per 100,000 — about three
times the peak of the previous winter — yet unmet need stood at 11.3%, *lower*
than the 12.4% observed during the far smaller December 2020 wave.

Two cautions apply. Excluding a period *because* it does not fit is a form of
post-hoc selection, so *r* = 0.72 should be read as evidence that **the
relationship is unstable across the period**, not as a better estimate of a single
underlying association. And even pre-Omicron the first-difference correlation
remains weak (*r* = 0.135, *p* = 0.51), so the co-movement is a matter of shared
levels and broad trend rather than synchronized wave-by-wave changes.

**Opinion:**
By winter 2021–22 vaccination was widespread, restrictions
were minimal, and widely-used home tests went largely unreported. A given case
count therefore corresponded to far less societal disruption, and far less
measured infection.

# Synthesis

Taken together, the two research questions show a weak but detectable signal between states and showcasing a correlated movement for the first two covid waves, which broke down at the omnicron infection surge. A possible explanation for the co-movement during the first two covid waves, could be, that the results of a surging infection rate would influence the policies regarding prevention of further outbreaks (social distance, working from home, ...). This pattern did not hold for the omnicron surge around early 2022, where restrictions became more minimal and vaccinations were widespread.

# Limitations

**Outcome construct.** "Needed counseling but did not get it" is a proxy, not a
direct measure of need: it conflates underlying need with barriers to access, and
over-represents people newly seeking support while excluding those already being
helped. Key determinants of mental-health need (healthcare capacity, socioeconomic composition ,...) are unobserved. Most importantly, high case rates coincided with lockdowns and social distancing,
which plausibly drove need themselves. Under that account the case rate proxies
policy stringency and isolation, making the association indirect rather than
causal.

**Missing data.** Estimates suppressed by NCHS for unreliability are dropped.
Suppression occurs *because* a subgroup sample was small, so the mechanism is
**not missing at random (MNAR)**. For our used data, 7 rows (0.42%) were missing.

Each observation is a whole state in one survey window, not a person. We therefore cannot tell whether the adults reporting
unmet need are the same adults affected by COVID.

