---
title: "COVID infection rates and unmet mental-health need - core analysis"
format: html
editor: visual
---

> Dataset background, licences, research questions and the full analysis plan are in [Project_Proposal.qmd](Project_Proposal.qmd). This file holds the runnable core analysis that answers the two research questions.

We want to know whether places affected harder by COVID also reported more unmet need for mental-health support. We measures one row per **US state** per **survey window**. The script below builds that table in three steps

-   the mental-health outcome
-   a comparable COVID case rate
-   join them and then answers the two research questions.

```{r setup}
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

# The outcome is a single switch so the whole analysis can be re-run for a
# different indicator (must match an Indicator value in the Household Pulse CSV).
OUTCOME_INDICATOR <- "Needed Counseling or Therapy But Did Not Get It, Last 4 Weeks"
```

## 1. Mental-health outcome: unmet need by state × survey window

The Household Pulse file stores every indicator/subgroup/state/period combination in one long table. We keep only the rows we need, the chosen outcome, split *By State*, with a real value, and rename the survey-window dates to short names we will reuse everywhere. We build the same thing once at state level (for RQ1) and once at national level (for RQ2).

```{r mh}
#| message: false
read_mh <- function(group, state_filter = NULL) {
  d <- read_csv(here::here("data", "raw", "Mental_Health_Care_in_the_Last_4_Weeks.csv"), show_col_types = FALSE) |>
    filter(Group == group, Indicator == OUTCOME_INDICATOR, !is.na(Value))
  if (!is.null(state_filter)) d <- filter(d, State == state_filter)
  d |>
    transmute(
      state      = State,
      win_start  = as.Date(`Time Period Start Date`, "%m/%d/%Y"),
      win_end    = as.Date(`Time Period End Date`,   "%m/%d/%Y"),
      unmet_need = Value                       # % of adults reporting the outcome
    )
}

mh     <- read_mh("By State")                          # one row per state × window
mh_nat <- read_mh("National Estimate", "United States") # one row per window

cat("state x window rows:", nrow(mh),
    "| states:", n_distinct(mh$state),
    "| windows:", n_distinct(mh$win_start), "\n")

cat("National Estimates:", nrow(mh_nat),
    "| windows:", n_distinct(mh$win_start), "\n")
```

//Todo: 51 \* 33 = 1683 but we habe 1676 obs. -\> there are a couple estimates Missing (see missing data)

## 2. COVID case rate per state, comparable across states

### What's in the two new datasets

Both new sources are small, official and tidy, so a short look is enough. We use only a few columns from each (the COVID file's `deaths` and `fips` go unused).

| File | Columns we use | Meaning |
|-------------------|---------------------------------|--------------------|
| `covid_us_states_nyt.csv` | `date`, `state`, `cases` | day · US state · **cumulative** confirmed cases |
| `us_state_population_2020.csv` | `state`, `population` | US state · 2020 resident population |

```{r new-data-overview}
#| message: false
pop       <- read_csv(here::here("data", "raw", "us_state_population_2020.csv"), show_col_types = FALSE)
covid_raw <- read_csv(here::here("data", "raw", "covid_us_states_nyt.csv"), show_col_types = FALSE)

glimpse(covid_raw)
glimpse(pop)

# Missing values in the columns we actually use (expect 0 in both):
covid_raw |> summarise(across(c(date, state, cases), ~ sum(is.na(.))))
pop       |> summarise(across(c(state, population),  ~ sum(is.na(.))))
```

**Why per 100k.** Raw case counts would just rank states by size (California will always look worst). To compare a small state with a large one we convert to **new cases per 100,000 residents**, using the population table above (US Census Bureau, 2020 Decennial Census, 50 states + DC).

The NYT file reports *cumulative* cases. The quantity we actually want is *new* cases on each day, which is just today's total minus yesterday's.

```{r covid}
#| message: false
covid_daily <- covid_raw |>
  arrange(state, date) |>
  mutate(new_cases = pmax(cases - lag(cases), 0), .by = state) |>  # cumulative -> daily
  inner_join(pop, by = "state") |>
  mutate(new_per100k = new_cases / population * 1e5)
```

## 3. Join COVID intensity into each survey window

Each survey window is a date range (about two weeks). For every state we add up its daily `new_per100k` that falls inside that window. The COVID change a state experienced while that survey was running. A **range join** (`between(date, win_start, win_end)`) matches each COVID day to the window it belongs to, so no manual looping is needed.

```{r join}
covid_win <- covid_daily |>
  inner_join(distinct(mh, state, win_start, win_end),
             by = join_by(state, between(date, win_start, win_end))) |>
  summarise(covid_rate = sum(new_per100k), .by = c(state, win_start, win_end))

joined <- inner_join(mh, covid_win, by = c("state", "win_start", "win_end"))

cat("joined rows:", nrow(joined), "\n")
summary(joined$covid_rate)
summary(joined$unmet_need)
```

## 4. RQ1: Does COVID intensity relate to unmet mental-health need?

With one comparable pair of numbers per state-window, we fit a straight line: does unmet need go up as the COVID case rate goes up? The slope answers the question; its 95% confidence interval and p-value tell us how sure we are.

```{r rq1-model}
m1 <- lm(unmet_need ~ covid_rate, data = joined)
summary(m1)

# Population claim -> report the coefficient with its 95% CI, not just the p-value.
cbind(estimate = coef(m1), confint(m1))
```

```{r rq1-plot}
ggplot(joined, aes(x = covid_rate, y = unmet_need)) +
  geom_point(alpha = 0.3, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "red") +
  labs(
    title = "Higher COVID case rates track slightly higher unmet mental-health need",
    subtitle = "Weak but significant positive association across US states and survey windows",
    x = "New COVID cases per 100k during the survey window",
    y = "% adults who needed counseling but did not get it"
  ) +
  theme_minimal()
```

## 5. RQ2: How did need and COVID co-move nationally over time?

RQ1 compares states; RQ2 instead follows the *whole country* through time. We build a national COVID case rate the same way (total new cases in each window ÷ US population) and plot it beside national unmet need on a shared time axis, so we can see whether the two rise and fall together across pandemic waves.

```{r rq2}
#| fig-cap: "Two-panel time series of national unmet mental-health need (top) and the national COVID-19 case rate (bottom), by survey window, Aug 2020 to Apr 2022. Plotting them on a shared time axis shows whether the two series rise and fall together across pandemic waves."
#| fig-alt: "Two stacked line charts sharing an x-axis of survey-window dates from Aug 2020 to Apr 2022. Top: percent of adults with unmet mental-health need. Bottom: new COVID cases per 100k. The COVID panel shows sharp winter peaks; the unmet-need panel is comparatively flat."
us_pop <- sum(pop$population)

covid_nat <- covid_daily |>
  inner_join(distinct(mh_nat, win_start, win_end),
             by = join_by(between(date, win_start, win_end))) |>
  summarise(covid_rate = sum(new_cases) / us_pop * 1e5, .by = c(win_start, win_end))

mh_nat |>
  left_join(covid_nat, by = c("win_start", "win_end")) |>
  select(win_start,
         `Unmet need (%)`    = unmet_need,
         `COVID cases /100k` = covid_rate) |>
  pivot_longer(-win_start) |>
  ggplot(aes(win_start, value)) +
  geom_line(colour = "darkblue") +
  geom_point(size = 1) +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(title = "National unmet mental-health need and COVID case rate over time",
       x = "Survey window start", y = NULL) +
  theme_minimal()
```

## Notes / next steps (for the remaining weeks)

-   RQ1 association is **positive and significant but weak** (see coefficient and R²). Refinements to try: a **time lag** (COVID surge → need a few weeks later) and **state fixed effects** to control for persistent between-state differences. These are presentation/robustness tasks, not new questions.
-   The outcome is not fixed yet. `OUTCOME_INDICATOR` swaps to any single Household Pulse indicator to repeat the whole analysis; we may also broaden "need" to the **overall movement across all four indicators** or each indicator's **average percentage change** between waves, and test which framing tracks COVID most clearly.
